### 2012-04-19 docs/mixture_models.tex
abstract and text tweaks

% Copyright 2012 David W. Hogg (NYU).
% All rights reserved.

@@

\begin{abstract}
Exponential and de~Vaucouleurs profiles are simple and
successful models for fitting two-dimensional images of galaxies.  One
numerical issue encountered in this kind of fitting is the pixel
rendering and convolution (or correlation) of the models with the
telescope point-spread function (PSF); these operations are slow, and
easy to get slightly wrong at small radii.  Here we exploit the
realization that these models can be approximated to arbitrary
accuracy with a mixture (linear sum) of two-dimensional Gaussians.
Mixtures of Gaussians are fast to render, fast to affine-transform,
and fast to convolve with mixture-of-Gaussian PSF models, all at
machine precision.  We present worked examples that can be directly
used in image fitting; we are using them ourselves.  We also advocate
modeling PSFs also as arbitrary mixtures of Gaussians.  Amusingly, in
the optically thin limit, a circularly symmetric mixture-of-Gaussian
two-dimensional model directly implies its own spherically symmetric
three-dimensional de-projection.
\end{abstract}

@@

In what follows we will exclusively use two-dimensional Gaussian or
Normal distributions, which look like
\begin{eqnarray}\displaystyle
\normal(\pos\given\mean,\var) &\equiv& \frac{1}{2\pi}\,\det(\var)^{-1/2}\,\exp(-\frac{1}{2}\,\transpose{[\pos-\mean]}\cdot\inverse{\var}\cdot[\pos-\mean])
\quad ,
\end{eqnarray}
where $\pos$ and $\mean$ are two-dimensional vectors (usually in the
focal plane or on the sky or something like that), and $\var$ is a
symmetric $2\times 2$ variance tensor or matrix, and implicitly the
vectors are column vectors.  A \emph{mixture of Gaussians} is a linear
superposition of Gaussians.  Any positive two-dimensional function
with finite support and finite total integral---including as a special
case any two-dimensional probability distribution function---can be
represented as a mixture of Gaussians to arbitrary accuracy (DSTN: It
would be nice to have a reference here).

We wish to make an approximation to the two-dimensional circular
exponential profile
\begin{eqnarray}\displaystyle
Q_{\exp}(\spos) &\approx& \frac{1}{Z_{\exp}}\,\exp(-\frac{|\spos|}{r_{\exp}})
\\
Z_{\exp} &=& XXXXX
\\
r_{\exp} &=& XXXXX
\quad ,
\end{eqnarray}
where $\spos$ is a dimensionless focal-plane position, and $Z_{\exp}$
and $r_{\exp}$ are parameters that ensure that the profile integrates
to unit flux and has unit half-light radius.  The position $\spos$ is
dimensionless because it parameterizes the unit-size and unit-flux
dimensionless function.  We seek the best (where ``best'' will be
defined below) $M_g$-Gaussian mixture approximation
\begin{eqnarray}\displaystyle
Q_{\exp}(\spos) &=& \sum_{m=1}^{M_g} g_m\,\normal(\spos\given\zero,\Gm_m)
\\
1 &=& \sum_{m=1}^{M_g} g_m
\\
\Gm_m &=& \sigma^2_{\exp,m}\,\identity
\quad ,
\end{eqnarray}
where all of the means are exactly zero and all of the variances
$\Gm_m$ in the mixture are scalar multiples of the identity matrix
$\identity$ because we are requiring this dimensionless function to be
precisely circular (so every component is itself circular and
concentric).  Similarly for the de~Vaucouleurs profile
\begin{eqnarray}\displaystyle
Q_{\dev}(\spos) &\approx& \frac{1}{Z_{\dev}}\,\exp(-\left[\frac{|\spos|}{r_{\dev}}\right]^{1/4})
\\
Z_{\dev} &=& XXXXX
\\
r_{\dev} &=& XXXXX
\\
Q_{\dev}(\spos) &=& \sum_{m=1}^{M_h} h_m\,\normal(\spos\given\zero,\Hm_m)
\\
1 &=& \sum_{m=1}^{M_h} h_m
\\
\Hm_m &=& \sigma^2_{\dev,m}\,\identity
\quad .
\end{eqnarray}

>>>>

% This file is part of the Tractor project.
% Copyright 2012 David W. Hogg (NYU) and Dustin Lang (Princeton).
% All rights reserved.

@@

\begin{abstract}
Exponential and de~Vaucouleurs profiles are simple and successful
models for fitting two-dimensional images of galaxies.  One numerical
issue encountered in this kind of fitting is the pixel rendering and
convolution (or correlation) of the models with the telescope
point-spread function (PSF); these operations are slow, and easy to
get slightly wrong at small radii.  Here we exploit the realization
that these models can be approximated to arbitrary accuracy with a
mixture (linear superposition) of two-dimensional Gaussians.  Mixtures
of Gaussians are fast to render and fast to affine-transform.  Most
importantly, if you have a mixture-of-Gaussian PSF model (which we
advocate), the PSF-convolved, affine-transformed galaxy models are
themselves mixtures of Gaussians and therefore very fast to compute
precisely.  We present worked examples that can be directly used in
image fitting; we are using them ourselves.  We discuss extensions to
Sersic models and applications of these ideas to three-dimensional
deprojection.
\end{abstract}

@@

...We need to put somewhere Lupton-like comments about sampling the
pixel-convolved-PSF-convolved scene...

In what follows we will make heavy use of two-dimensional Gaussian or
Normal distributions, which look like
\begin{eqnarray}\displaystyle
\normal(\pos\given\mean,\var) &\equiv& \frac{1}{2\pi}\,\det(\var)^{-1/2}\,\exp(-\frac{1}{2}\,\transpose{[\pos-\mean]}\cdot\inverse{\var}\cdot[\pos-\mean])
\quad ,
\end{eqnarray}
where $\pos$ and $\mean$ are two-dimensional vectors (usually in the
focal plane or on the sky or something like that), and $\var$ is a
symmetric $2\times 2$ variance tensor or matrix, and implicitly the
vectors are column vectors.  A \emph{mixture of Gaussians} is a linear
superposition of Gaussians.  Any positive two-dimensional function
with finite support and finite total integral---including as a special
case any two-dimensional probability distribution function---can be
represented as a mixture of Gaussians to arbitrary accuracy (DSTN: It
would be nice to have a reference here).

We wish to make an approximation $Q_{\exp}(\cdot)$ to the
two-dimensional circular exponential profile
\begin{eqnarray}\displaystyle
Q_{\exp}(\spos) &\approx& \frac{1}{Z_{\exp}}\,\exp(-\frac{|\spos|}{r_{\exp}})
\\
Z_{\exp} &=& X.XXXX
\\
r_{\exp} &=& X.XXXX
\quad ,
\end{eqnarray}
where $\spos$ is a dimensionless focal-plane position, and $Z_{\exp}$
and $r_{\exp}$ are parameters that ensure that the profile integrates
to unit flux and has unit half-light radius.  The position $\spos$ is
dimensionless because it parameterizes the unit-size and unit-flux
dimensionless function.  We seek the best (where ``best'' will be
defined below) $M_g$-Gaussian mixture approximation
\begin{eqnarray}\displaystyle
Q_{\exp}(\spos) &=& \sum_{m=1}^{M_g} g_m\,\normal(\spos\given\zero,\Gm_m)
\\
1 &=& \sum_{m=1}^{M_g} g_m
\\
\Gm_m &=& \sigma^2_{\exp,m}\,\identity
\quad ,
\end{eqnarray}
where all of the means are exactly zero and all of the variances
$\Gm_m$ in the mixture are scalar multiples of the identity matrix
$\identity$ because we are requiring this dimensionless function to be
precisely circular (so every component is itself circular and
concentric).  Similarly for the de~Vaucouleurs profile
\begin{eqnarray}\displaystyle
Q_{\dev}(\spos) &\approx& \frac{1}{Z_{\dev}}\,\exp(-\left[\frac{|\spos|}{r_{\dev}}\right]^{1/4})
\\
Z_{\dev} &=& X.XXXX
\\
r_{\dev} &=& X.XXXX
\\
Q_{\dev}(\spos) &=& \sum_{m=1}^{M_h} h_m\,\normal(\spos\given\zero,\Hm_m)
\\
1 &=& \sum_{m=1}^{M_h} h_m
\\
\Hm_m &=& \sigma^2_{\dev,m}\,\identity
\quad .
\end{eqnarray}
